Discrete Quantum Theories and Computing

### 2018-12-29 dissertation.tex
dissertation.lyx "Chapter 1 Introduction" Updating...

After first suggested by Richard Feynman 40 years ago~\cite{Feynman1982Simulating},
IBM, \textitalian{Rigetti}, Google, Microsoft, and many other companies
and countries start to build realistic quantum computers recently
\cite{Shieber2018,Kahn2019}, and seek more efficient ways to simulate
chemical molecules and to break down RSA \cite{boneh1999twenty,wiki:RSAFactoring}.
No matter how to, classical or quantum, attack these two tasks, they
require two different types of quantities. On one hand, a question
and its answer of RSA are both discrete integers. On the other hand,
molecules are described by quantities including the domain of their
position wave functions and the probability amplitudes of these wave
functions, which look like continuous at first sight. However, we
want to argue that either these quantities might later be discovered
to be discrete, or the exact values of these continuous quantities
may never be pinned down. Hence, modeling quantum computing based
on the exact values of continuous quantities cannot really describe
the computational power of a realistic quantum computer. To better
understand this issue without struggling with the quantum theory,
it is instructive to review how classical quantities are characterized
and computed.

@@

Although it is convenient to assume physical quantities is as continuous
as their mathematical models, each physical quantity discussed in
the previous paragraph could not have a simple one-to-one correspondence
with a real number. Chuang-tzu's stick are made by molecules and cannot
be cut in halves endlessly. The input charge of an integrator must
actually be an integer multiple of the elementary charge. A prefect
square cannot be draw physically. Even we don't draw it, but merely
ask whether it exists in the space. This question is still beyond
our ability to answer because whether the space itself is ultimately
discrete or continuous is still an unsolved question. Even if a prefect
square exists physically, we may not able to decide whether the ``square''
we discussed is really prefect or almost prefect differing only by
an arbitrarily small extent. Similar situation applying to the length
of a slide rule, it is beyond our ability to decide whether its length
is ultimately discrete or continuous. Even if its length is continuous,
identifying which real number represents its length requires infinite
precision, but the precision in reality is limited generally three
or four significant figures~\cite{wiki:AnalogComputer}. As the results,
even if analog computers really store real numbers, they cannot be
precisely read, written, and used to branch the computation. The opposite
of the last statement is sometimes assumed by real computational models.
For example, Blum, Cucker, Shub, and Smale modeled their BCSS machine
to allow the operations of deciding whether a number is greater or
equal to zero over real numbers and whether a number is exactly zero
over complex numbers \cite{BSS1989,Ziegler2007,blum2012complexity}.
Hence, when the quantities used to branch the computation is closed
to zero, the branch chosen by a BCSS machine cannot reliably predict
the branch chosen by a realistic analog computer. Since two branches
of computations may not have any relations, the follow-up computational
paths may be significantly different, and may not be compensated by
error analysis techniques easily like the butterfly effect. Therefore,
the computational power of analog computers might be far weaker than
the prediction based on this kind of theoretical models. \todo{Maybe
extended to consider probabilistic model? }

\begin{sidewaystable}
\caption{\label{tab:Comparison}Comparison between the classical quantities
and their mathematical models with the inspired quantum models. \todo{Center
entries vertically.}}

\centering{}\centering{}%
\begin{tabularx}{\linewidth}{ccYY}
\toprule 
\addlinespace
Physical quantities & Discrete & Discrete with extreme small unit & Continuous?\tabularnewline\addlinespace
\addlinespace
\addlinespace
Computational devices & Digital computer & Analog computer & Analog computer\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Mathematical representation & Integers & Real numbers & Real numbers\tabularnewline\addlinespace
\addlinespace
\addlinespace
Computational model  & Turing machine & BCSS machine & BCSS machine\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
How the models realize? & Reliably & 
Not reliably, but after we have a better technology to manipulate the minimum units, we can predict the physical phenomenon and their computational power by a better discrete model. & 
Not reliably, because precision can never be high enough, and the difference between their computational power might not be compensated by error analysis techniques. \tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Inspired quantum models &  & Quantum Theories and Computing over Finite Fields & Quantum Interval-Valued Probability Measures (QIVPM)\tabularnewline\addlinespace
\addlinespace
\bottomrule
\end{tabularx}
\end{sidewaystable}
The situations for classical quantities and the computational power
above them are summarized in Table~\ref{tab:Comparison}, and two
possibilities dealing with real-valued models inspires us two types
of quantum models. Similar to classical analog computers, the quantum
circuit model, the most widely used model of quantum computing, manipulates
quantities which are not discrete apparently: the probability amplitudes
of quantum states. To address the similar inconsistency, we first
review the ``conventional quantum theory''\footnote{Alternative terminology in the literature includes ``actual,'' ``standard,''
and ``ordinary'' quantum theory.} in Chapter~\ref{chap:Conventional-Quantum-Theory}, where the probability
amplitudes are in the field of complex numbers. If the probability
amplitudes are assumed to be ultimately discrete, we replace the field
of complex numbers by discrete finite fields and consider the quantum
theories and computing over finite fields. Another possibilities could
be that the probability amplitudes are never quantized no matter how
hard we try, and we could never access their exactly values. Notice
that the probability amplitudes can never be accessed directly in
an experiment, the accessible properties of a quantum system is its
measurement results and the probability of an event. Hence, to incorporate
the finite precision measurement into the quantum theory, we consider
a framework called quantum interval-valued probability measures (QIVPM).

\begin{comment}
Our first discrete model replaces the complex numbers by a unrestricted
finite field~$\mathbb{F}_{q}$ \cite{Schumacher2012-SCHMQT,DQT2014,SchumacherWestmoreland2010}.
In this model, an $n$-qubit state is a non-zero vector in $\mathbb{F}_{q}^{2^{n}}$.
When we measure a given state~$\ket{\Phi}$, an observable is replaced
by a basis~$\mathcal{B}$ of $\mathbb{F}_{q}^{2^{n}}$. Since $\ket{\Phi}$
can be represented by the summation across a unique subset~$\mathcal{S}$
of the basis~$\mathcal{B}$, whether it is possible or impossible
to measure a basis vector~$\ket{i}$ depends on whether $\ket{i}$
is in $\mathcal{S}$ or not. Because this model only predicts whether
a result is possible or impossible, it is called the modal quantum
theory. Its computational model, called the modal quantum computing,
is far from conventional quantum computing. Although modal quantum
computing can express simple algorithms such as quantum teleportation
\cite{BennettBrassardEtAl1993,peres1995quantum,Mermin2007,Jaeger2007},
it is so weak that it cannot express Deutsch's algorithm \cite{Deutsch1985,Mermin2007}.
This quantum theory is, however, also so powerful that it can be used
to solve an unstructured database search of size $N$ using $O(\log(N))$
steps, which outperforms the known asymptotic bound $O(\sqrt{N})$
in conventional quantum computing \cite{Grover:1996:FQM:237814.237866,BennettBernsteinBrassardVazirani1997,Mermin2007,Jaeger2007}.
\end{comment}
When replacing the continuum of complex numbers by discrete finite
fields, we first doesn't restrict our quantum theory and computing
over which finite fields in Chapter~\ref{chap:Unrestricted-Finite-Fields}
\cite{Schumacher2012-SCHMQT,DQT2014,SchumacherWestmoreland2010}.
This theory is called modal quantum theory because it is so weak that
it only predict whether the measure result is possible or impossible,
and even the expectation value of an observable is not able to be
defined. Although a version of quantum theory defined over a two-valued
field can express simple algorithms such as quantum teleportation
\cite{BennettBrassardEtAl1993,peres1995quantum,Mermin2007,Jaeger2007},
it is so weak that it cannot express Deutsch's algorithm \cite{Deutsch1985,Mermin2007}.
This quantum theory is, however, also so powerful that it can be used
to solve the $\usat$ problem, which is as hard as a general $\NP$-complete
problem.

\todo{Introduction Cleaned up until here. }Next, in Sec.~\ref{discretequantumtheoryI}
to \ref{sec:Geometry-of-Entangled-States}, we improve on this by
showing that for finite fields of order $p^{2}$, with the prime $p$
of the form $4\ell+3$ ($\ell$ a non-negative integer), the complex
numbers have extremely compelling and natural discrete analogs that
permit a great many of the standard requirements of quantum computing
to be preserved. Under suitable conditions, we have amplitude-based
partitions of unity, unitary transformations, and entanglement, as
well as solutions to deterministic quantum algorithms such as the
algorithms of Deutsch, Simon, and Bernstein-Vazirani, though still
with some bothersome shortcomings. Because of the modular nature of
arithmetic in the finite complex field, it is not possible to define
an inner product in the usual sense, and we show in Sec.~\ref{subsec:Partial-Algorithm}
that this leads to excessive computational power for the unstructured
database search problem for certain database sizes.

@@

Finally, we shift our attention to consider quantum interval-valued
probability measures (IVPMs), which potentially embody both finite
precision measurement and a sensible correspondence to standard quantum
probability. This interval-valued framework not only provides a natural
generalization of both classical IVPMs and conventional quantum probability
measures, but also allows us to establish bounds on the validity of
the Kochen-Specker and Gleason theorems in realistic experimental
environments.

@@

\begin{tabularx}{\linewidth}{YYYYYYY}
\toprule 
\addlinespace
 & Conventional & Modal & Discrete (I) & Discrete (II) & QPMFF & QIVPM\tabularnewline\addlinespace
\addlinespace
\midrule
\midrule 
\addlinespace
 & Chapter~\ref{chap:Conventional-Quantum-Theory} & Sec.~\ref{modalquantum} & Sec.~\ref{discretequantumtheoryI} to \ref{sec:Geometry-of-Entangled-States} & Sec.~\ref{sec:Discrete-Quantum-Theory-(II)} & Sec.~\ref{sec:Toward-IVPM} & Chapter~\ref{chap:QIVPM}\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
States space & $\mathbb{C}^{D}$ & $\Fq^{D}$ & $\Fpp^{D}$ & Local region in $\Fpp^{D}$ & $\Fpp^{D}$ & $\mathbb{C}^{D}$\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Likelihood of events is predicted by & Real-valued probability & Possible or impossible & Possible or impossible & Cardinal probability & Real-valued probability, but no sensible Born rule & Interval-valued probability\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Expectation value & Defined & Undefined & Undefined & Undefined &  & Defined\tabularnewline\addlinespace
\addlinespace
\midrule
\midrule 
\addlinespace
 &  & Sec.~\ref{modalquantumcomputing} & Sec.~\ref{discretequantumcomputingI} & Sec.~\ref{discretequantumcomputingII} &  & \tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Deutsch's algorithm & Yes & Maybe no & Yes & Yes &  & \tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Solve $\usat$ efficiently & Unlikely & Yes & Partially & Unlikely &  & \tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Grover search algorithm & Yes &  &  & Yes &  & \tabularnewline\addlinespace
\addlinespace
\bottomrule
\end{tabularx}
\end{sidewaystable}

@@

In this section we want to build a theory of discrete vector spaces
that approximates as closely as possible the features of conventional
quantum theory. Such a structure would ideally consist of the following:
(i) a vector space over the field of complex numbers, and (ii) an
inner product $\ip{\Phi}{\Psi}$ associating to each pair of vectors
a complex number, and satisfying the following properties: 
\begin{enumerate}[label=(\Alph{enumi})]
\item \label{enu:inner-product-complex-conjugate}$\ip{\Phi}{\Psi}$ is
the complex conjugate of $\ip{\Psi}{\Phi}$; 
\item \label{enu:inner-product-linear}$\ip{\Phi}{\Psi}$ is conjugate linear
in its first argument and linear in its second argument; 
\item \label{enu:inner-product-positive-definite}$\ip{\Psi}{\Psi}$ is
always non-negative and is equal to~0 only if $\ket{\Psi}$ is the
zero vector. 
\end{enumerate}
It turns out that a vector space defined over a finite field cannot
have an inner product satisfying the properties above. However, we
will introduce an Hermitian ``dot product'' satisfying some of those
properties.

@@

\begin{definition}[Hermitian dot product] Given vectors $\ket{\Phi}$
and $\ket{\Psi}\in\mathcal{H}$, it can be shown~\cite{grove2002classical,Wan2006}
the Hermitian dot product is always reducible to the form 
\begin{equation}
\ip{\Phi}{\Psi}=\sum_{i=0}^{D-1}\beta_{i}^{p}\alpha_{i}\,.\label{innerprod}
\end{equation}
\end{definition} Two vectors $\ket{\Phi}$ and $\ket{\Psi}\in\mathcal{H}$
are said to be orthogonal if $\ip{\Phi}{\Psi}=0$. This product satisfies
conditions \ref{enu:inner-product-complex-conjugate} and \ref{enu:inner-product-linear}
for inner products but violates condition \ref{enu:inner-product-positive-definite}
since in every finite field there always exists a non-zero vector
$\ket{\Psi}$ such that $\ip{\Psi}{\Psi}=0$. The reason is that addition
in finite fields eventually ``wraps around'' (because of their cyclic
or modular structure), allowing the sum of non-zero elements to be
zero. The fraction of non-zero vectors satisfying $\ip{\Psi}{\Psi}=0$
decreases with the order $p$.

>>>>

After first suggested by Richard Feynman 40 years ago~\cite{Feynman1982Simulating},
IBM, \textitalian{Rigetti}, Google, Microsoft, and many other companies
and countries start to build realistic quantum computers recently
\cite{Shieber2018,Kahn2019}, and seek more efficient ways to simulate
chemical molecules and to break down RSA \cite{boneh1999twenty,wiki:RSAFactoring}.
No matter how to, classical or quantum, attack these two tasks, they
require two different types of quantities. On one hand, a question
and its answer of RSA are both discrete integers. On the other hand,
molecules are described by quantities including the domain of their
position wave functions and the probability amplitudes of these wave
functions, which look like continuous at first sight. However, either
these quantities might later be discovered to be discrete, or their
exact values may never be pinned down. Hence, modeling quantum computing
based on the exact values of continuous quantities cannot really describe
the computational power of a realistic quantum computer. To better
understand this issue without struggling with the quantum theory,
it is instructive to review how classical quantities are characterized
and computed.

@@

Although it is convenient to assume physical quantities is as continuous
as their mathematical models, each physical quantity discussed in
the previous paragraph could not have a simple one-to-one correspondence
with a real number. Chuang-tzu's stick are made by molecules and cannot
be cut in halves endlessly. The input charge of an integrator must
actually be an integer multiple of the elementary charge. A prefect
square cannot be draw physically. Even we don't draw it, but merely
ask whether it exists in the space. This question is still beyond
our ability to answer because whether the space itself is ultimately
discrete or continuous is still an unsolved question. Even if a prefect
square exists physically, we may not able to decide whether the ``square''
we discussed is really prefect or almost prefect differing only by
an arbitrarily small extent. Similar situation applying to the length
of a slide rule, it is beyond our ability to decide whether its length
is ultimately discrete or continuous. Even if its length is continuous,
identifying which real number represents its length requires infinite
precision, but the precision in reality is limited generally three
or four significant figures~\cite{wiki:AnalogComputer}. As the results,
even if analog computers really store real numbers, they cannot be
precisely read, written, and used to branch the computation. The opposite
of the last statement is sometimes assumed by real computational models.
For example, Blum, Cucker, Shub, and Smale modeled their BCSS machine
to allow the operations of deciding whether a number is greater or
equal to zero over real numbers and whether a number is exactly zero
over complex numbers \cite{BSS1989,Ziegler2007,blum2012complexity}.
Hence, when the quantities used to branch the computation is closed
to zero, the branch chosen by a BCSS machine cannot reliably predict
the branch chosen by a realistic analog computer. Since two branches
of computations may not have any relations, the follow-up computational
paths may be significantly different, and these difference may not
be compensated by error analysis techniques easily like the butterfly
effect. Therefore, the computational power of analog computers might
be far weaker than the prediction based on this kind of theoretical
models.%
\begin{comment}
\todo{Maybe extended to consider probabilistic model? }
\end{comment}

\begin{sidewaystable}
\caption{\label{tab:Comparison}Comparison between the classical quantities
and their mathematical models with the inspired quantum models.}

\centering{}\centering{}%
\begin{tabularx}{\linewidth}{ccYY}
\toprule 
\addlinespace
Physical quantities & Discrete & Discrete with extreme small unit & Continuous?\tabularnewline\addlinespace
\addlinespace
\addlinespace
Computational devices & Digital computer & Analog computer & Analog computer\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Mathematical representation & Integers & Real numbers & Real numbers\tabularnewline\addlinespace
\addlinespace
\addlinespace
Computational model  & Turing machine & BCSS machine & BCSS machine\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
How the models realize? & Reliably & 
Not reliably, but after we have a better technology to manipulate the minimum units, we can predict the physical phenomenon and their computational power by a better discrete model. & 
Not reliably, because precision can never be high enough, and the difference between their computational power might not be able compensated by error analysis techniques. \tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Inspired quantum models &  & Quantum Theories and Computing over Finite Fields & Quantum Interval-Valued Probability Measures (QIVPM)\tabularnewline\addlinespace
\addlinespace
\bottomrule
\end{tabularx}
\end{sidewaystable}
The situations for classical quantities and the computational power
above them are summarized in Table~\ref{tab:Comparison}, and the
two possibilities dealing with real-valued models inspires us two
types of quantum models. Similar to classical analog computers, the
quantum circuit model, the most widely used model of quantum computing,
manipulates quantities which are not discrete apparently: the probability
amplitudes of quantum states. To address the similar inconsistency,
we first review the ``conventional quantum theory''\footnote{Alternative terminology in the literature includes ``actual,'' ``standard,''
and ``ordinary'' quantum theory.} in Chapter~\ref{chap:Conventional-Quantum-Theory}, where the probability
amplitudes are in the field of complex numbers. If the probability
amplitudes are assumed to be ultimately discrete, we replace the field
of complex numbers by discrete finite fields and consider the quantum
theories and computing over finite fields in Chapter~\ref{chap:Unrestricted-Finite-Fields}
and \ref{chap:DQT=000026DQC}. Another possibilities could be that
the probability amplitudes would never be quantized no matter how
hard we try. Notice that the probability amplitudes can never be accessed
directly in an experiment, the accessible properties of a quantum
system is its measurement results and the probability of an event.
Hence, to incorporate the finite precision measurement into the quantum
theory, we consider a framework called quantum interval-valued probability
measures (QIVPM) in Chapter~\ref{chap:QIVPM}.

\begin{comment}
After a review of finite fields in Section~\ref{sec:background},
we proceed with a sequence of finite-field approaches that lead more
and more closely to the properties of conventional quantum computing.
\end{comment}
\begin{comment}
An $n$-qubit pure state in the conventional quantum theory (CQT)
is represented as a vector in the $2^{n}$-dimensional Hilbert space.
If we eliminate all symmetries, an irreducible state is actually a
point in the projective Hilbert space \cite{MosseriDandoloff2001,Jaeger2007}
also known as the complex projective space~$\CP{2^{n}-1}$ \cite{Hatcher2001,Bengtsson2007}.
These irreducible quantum states can also be classified as product
states and entangled states, and the latter one plays the essential
role for pure-state algorithms \cite{Mermin2007,Jaeger2007}.

Given a pure state~$\ket{\Phi}$, when we measure an observable represented
by an Hermitian matrix~$\mathbf{O}$, the measurement result is one
of the eigenvalues of $\mathbf{O}$. The probability of getting a
particular eigenvalue~$\lambda$ is $\melem{\Phi}{P}{\Phi}$, where
$P$ is the projection operator onto the eigenspace of $\lambda$.
This rule of computing the probability is called the Born rule \cite{Born1983,Mermin2007,Jaeger2007},
which is used when we want to extract information from a quantum computer.
For any mixed state~$\rho$, the generalized Born rule induces a
conventional quantum probability measure $\muB_{\rho}:\events\rightarrow[0,1]$,
where $\events$ is the set of all projection operators on a given
Hilbert space. Conversely, any quantum probability measure $\mu:\events\rightarrow[0,1]$
can be induced from a mixed state $\rho$ in the Hilbert space of
dimension $D\ge3$ according to Gleason's theorem \cite{gleason1957,Redhead1987-REDINA,peres1995quantum,RichmanBridges1999,Hamhalter2013}.
In another word, this state $\rho$ is the unique state consistent
with any given quantum probability measure.
\end{comment}
The review of CQT in Chapter~\ref{chap:Conventional-Quantum-Theory}
starts from the geometrical structure of states in Sec.~\ref{CQC1qubitBloch.sec},
where a pure state is represented as a vector in the Hilbert space,
and an irreducible state is actually a point in the projective Hilbert
space \cite{MosseriDandoloff2001,Jaeger2007} also known as the complex
projective space \cite{Hatcher2001,Bengtsson2007}. In Sec.~\ref{sec:fuzzy},
we will construct a quantum probability space based on classical probability
space, apply Gleason's theorem \cite{gleason1957,Redhead1987-REDINA,peres1995quantum,RichmanBridges1999,Hamhalter2013}
to recover the Born rule \cite{Born1983,Mermin2007,Jaeger2007}, and
define the expectation value of an observable. These expectation values
will be used to defined purity which provides an easy way to decide
whether a state is entangled or not and to defined the maximum entangled
states.

\begin{comment}
In Section~\ref{modalquantum}, we examine previously-introduced
quantum theories defined over unrestricted finite fields and show
in Section~\ref{modalquantumcomputing} that this approach leads
to theories with such bizarre powers that they are probably unphysical.
Although a version of quantum theory defined over a two-valued field
can express simple algorithms such as quantum teleportation, it is
so weak that it cannot express Deutsch's algorithm. This quantum theory
is, however, also so powerful that it can be used to solve an unstructured
database search of size $N$ using $O(\log(N))$ steps, which outperforms
the known asymptotic bound $O(\sqrt{N})$ in conventional quantum
computing.
\end{comment}
\begin{comment}
Our first discrete model replaces the complex numbers by a unrestricted
finite field~$\mathbb{F}_{q}$ \cite{Schumacher2012-SCHMQT,DQT2014,SchumacherWestmoreland2010}.
In this model, an $n$-qubit state is a non-zero vector in $\mathbb{F}_{q}^{2^{n}}$.
When we measure a given state~$\ket{\Phi}$, an observable is replaced
by a basis~$\mathcal{B}$ of $\mathbb{F}_{q}^{2^{n}}$. Since $\ket{\Phi}$
can be represented by the summation across a unique subset~$\mathcal{S}$
of the basis~$\mathcal{B}$, whether it is possible or impossible
to measure a basis vector~$\ket{i}$ depends on whether $\ket{i}$
is in $\mathcal{S}$ or not. Because this model only predicts whether
a result is possible or impossible, it is called the modal quantum
theory. Its computational model, called the modal quantum computing,
is far from conventional quantum computing. Although modal quantum
computing can express simple algorithms such as quantum teleportation
\cite{BennettBrassardEtAl1993,peres1995quantum,Mermin2007,Jaeger2007},
it is so weak that it cannot express Deutsch's algorithm \cite{Deutsch1985,Mermin2007}.
This quantum theory is, however, also so powerful that it can be used
to solve an unstructured database search of size $N$ using $O(\log(N))$
steps, which outperforms the known asymptotic bound $O(\sqrt{N})$
in conventional quantum computing \cite{Grover:1996:FQM:237814.237866,BennettBernsteinBrassardVazirani1997,Mermin2007,Jaeger2007}.
\end{comment}
When replacing the continuum of complex numbers by discrete finite
fields, we first doesn't restrict the chosen finite fields in Chapter~\ref{chap:Unrestricted-Finite-Fields}
\cite{Schumacher2012-SCHMQT,DQT2014,SchumacherWestmoreland2010}.
This theory is called modal quantum theory because it can only predict
whether a measure result is possible or impossible, but not its probability.
Although a version of quantum theory defined over the two-valued field
can express simple algorithms such as quantum teleportation \cite{BennettBrassardEtAl1993,peres1995quantum,Mermin2007,Jaeger2007},
it is so weak that it cannot express Deutsch's algorithm \cite{Deutsch1985,Mermin2007}.
This quantum theory is, however, also so powerful that it can be used
to solve the $\usat$ problem~\cite{Papadimitriou1993}, which is
as hard as a general $\NP$-complete problem~\cite{Valiant198685}.

\begin{comment}
Next, in Section \ref{discretequantumtheoryI}, we improve on this
by showing that for finite fields of order $p^{2}$, with the prime
$p$ of the form $4\ell+3$ ($\ell$ a non-negative integer), the
complex numbers have extremely compelling and natural discrete analogs
that permit a great many of the standard requirements of quantum computing
to be preserved. Under suitable conditions, we have amplitude-based
partitions of unity, unitary transformations, and entanglement, as
well as solutions to deterministic quantum algorithms such as the
algorithms of Deutsch, Simon, and Bernstein-Vazirani~\cite{NCbook,Mermin},
though still with some bothersome shortcomings. Because of the modular
nature of arithmetic in the finite complex field, it is not possible
to define an inner product in the usual sense, and we show in Section~\ref{discretequantumcomputingI}
that this leads to excessive computational power for the unstructured
database search problem for certain database sizes.
\end{comment}
\begin{comment}
Our second model called discrete quantum theory (I) considers only
finite fields of order $p^{2}$, with the prime $p$ of the form $4\ell+3$
($\ell$ a non-negative integer) \cite{geometry2013,DQT2014}. In
this model, an irreducible state is a vector in $\mathbb{F}_{p^{2}}^{2^{n}}$,
and can be reduced to a point in the discrete complex projective space~$\mathbb{DCP}^{2^{n}-1}$.
Among these irreducible states, the product states and entangled states
can not only be identified as in CQT but also be counted for different
$p$. Although the state space is a natural discrete analog to CQT,
this model still only predicts whether a measurement result is possible
or impossible, because it is not possible to define an inner product
in the usual sense due to the modular arithmetic~\cite{grove2002classical}.
Under suitable conditions, we can have deterministic quantum algorithms
such as the algorithms of Deutsch, Simon \cite{Simon:1994:PQC:1398518.1399019,Mermin2007,Jaeger2007},
and Bernstein-Vazirani \cite{Bernstein:1993:QCT:167088.167097,Mermin2007},
but this still leads to excessive computational power for the unstructured
database search problem for certain database sizes.
\end{comment}
Next, in Sec.~\ref{discretequantumtheoryI} to \ref{sec:Geometry-of-Entangled-States},
we improve on this by showing that for finite fields of order $p^{2}$,
$\Fpp$, with the prime $p$ of the form $4\ell+3$ ($\ell$ a non-negative
integer), the complex numbers have extremely compelling and natural
discrete analogs that permit a great many of the standard requirements
of quantum computing to be preserved. In this model, the states of
a discrete $d$-dimensional system is a vector in the $d$-dimensional
vector space over $\Fpp$. \todo{Introduction Cleaned up until here.
} Because the states of a discrete $n$-qubit system are in principle
enumerable, we can count the number of states, and determine the proportions
of entangled and unentangled states. Under suitable conditions, we
have amplitude-based partitions of unity, unitary transformations,
and entanglement, as well as solutions to deterministic quantum algorithms
such as the algorithms of Deutsch, Simon \cite{Simon:1994:PQC:1398518.1399019,Mermin2007,Jaeger2007},
and Bernstein-Vazirani \cite{Bernstein:1993:QCT:167088.167097,Mermin2007},
though still with some bothersome shortcomings. Because of the modular
nature of arithmetic in the finite complex field, it is not possible
to define an inner product in the usual sense, and we show in Sec.~\ref{subsec:Partial-Algorithm}
that this leads to excessive computational power for the unstructured
database search problem for certain database sizes.

@@

\begin{comment}
We are led, in Sections~\ref{discretequantumtheoryIIa} and~\ref{discretequantumtheoryIIb},
to develop a framework with further restrictions on~$p$ that \emph{locally\/}
recovers the structure and expected properties of conventional quantum
theory. Section~\ref{discretequantumtheoryIIa} locally recovers
the inner product space and Section~\ref{discretequantumtheoryIIb}
locally recovers a notion of probability. The development in both
sections exploits the fact that longer sequences of \textit{ordered\/}
numbers appear in the quadratic residues (numbers with square-roots
in the field) as the size of the field increases. Discrete quantum
computations whose calculations are confined to numbers in this ordered
sequence resemble conventional quantum computations. The size of the
field $p$ plays an important role in describing the resources needed
for the computation as larger problem sizes require a larger field
size to represent all intermediate numerical values. A significant
feature of our framework is that the resources needed for the measurement
process can be separated from the resources needed by the evolution
of the system being modeled. This interplay between the resources
used by the system under study and the resources used for the observation
process is a significant concept that is nonexistent in conventional
quantum computing and is exposed by our careful accounting of resources.

We note that the conventional mathematical framework based on the
real numbers allows one to distinguish states whose measurement outcomes
differ by infinitesimally small probabilities, e.g., $10^{-100}$
vs.~$0$. In the proposed framework of discrete quantum computing,
the finite size of the field implies a maximum precision for measurement:
a ``small'' field represents limited resources with which it becomes
impossible to distinguish states whose measurement outcomes differ
by an amount less than the resolution afforded by the field. It is
possible, however, to discriminate between such states at the cost
of moving to a larger field, i.e., by investing more resources in
the measurement process. We formalize this approach to measurement
using the novel notion of \textit{cardinal probability\/}, with numerical
labels corresponding to ``more probable, the same, or less probable,''
rather than a percentage-based likelihood measure. In cardinal probability,
relative outcomes are associated with intervals of ambiguity that
get smaller and more precise as the size of the field increases.

Finally, in Section~\ref{discretequantumcomputingII}, we apply our
discrete quantum theory to the study of two representative algorithms,
the deterministic Deutsch-Jozsa algorithm and the probabilistic Grover
algorithm~\cite{NCbook,Mermin}. The first algorithm highlights the
role played by the size of the field $p$ in determining the actual
resources required for computation as the number of input bits $n$
increases, a concept nonexistent in conventional quantum computing.
The second algorithm highlights, in addition, the dependence of the
precision of measurement (via cardinal probabilities) on the size
of the field, another nonexistent concept in conventional quantum
computing.
\end{comment}
\begin{comment}
Our third model, discrete quantum theory (II)~\cite{DQT2014}, restricts
states in some local region of $\mathbb{F}_{p^{2}}^{2^{n}}$. Within
a local region, a notion of inner product and probability could be
recovered. Its discrete quantum computing can be further applied to
the deterministic Deutsch-Jozsa algorithm \cite{DeutschJozsa1992,Jaeger2007}
and the probabilistic Grover algorithm \cite{Grover:1996:FQM:237814.237866,Mermin2007,Jaeger2007}.
\end{comment}
Finally, we shift our attention to consider quantum interval-valued
probability measures (IVPMs), which potentially embody both finite
precision measurement and a sensible correspondence to standard quantum
probability. This interval-valued framework not only provides a natural
generalization of both classical IVPMs and conventional quantum probability
measures, but also allows us to establish bounds on the validity of
the Kochen-Specker and Gleason theorems in realistic experimental
environments.

@@

\begin{tabularx}{\linewidth}{YYYYYYY}
\toprule 
\addlinespace
 & Conventional & Modal & Discrete (I) & Discrete (II) & QPMFF & QIVPM\tabularnewline\addlinespace
\addlinespace
\midrule
\midrule 
\addlinespace
 & Chapter~\ref{chap:Conventional-Quantum-Theory} & Sec.~\ref{modalquantum} & Sec.~\ref{discretequantumtheoryI} to \ref{sec:Geometry-of-Entangled-States} & Sec.~\ref{sec:Discrete-Quantum-Theory-(II)} & Sec.~\ref{sec:Toward-IVPM} & Chapter~\ref{chap:QIVPM}\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
States space & $\mathbb{C}^{D}$ & $\Fq^{D}$ & $\Fpp^{D}$ & Local region in $\Fpp^{D}$ & $\Fpp^{D}$ & $\mathbb{C}^{D}$\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Likelihood of events is predicted by & Real-valued probability & Possible or impossible & Possible or impossible & Cardinal probability & Real-valued probability, but no sensible Born rule & Interval-valued probability\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Expectation value & Defined & Undefined & Formally defined & Undefined &  & Defined\tabularnewline\addlinespace
\addlinespace
\midrule
\midrule 
\addlinespace
 &  & Sec.~\ref{modalquantumcomputing} & Sec.~\ref{discretequantumcomputingI} & Sec.~\ref{discretequantumcomputingII} &  & \tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Deutsch's algorithm & Yes & Maybe no & Yes & Yes &  & \tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Solve $\usat$ efficiently & Unlikely & Yes & Partially & Unlikely &  & \tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Grover search algorithm & Yes &  &  & Yes &  & \tabularnewline\addlinespace
\addlinespace
\bottomrule
\end{tabularx}
\end{sidewaystable}

@@

In this section we want to build a theory of discrete vector spaces
that approximates as closely as possible the features of conventional
quantum theory. Such a structure would ideally consist of the following:
(i) a vector space over the field of complex numbers, and (ii) an
inner product $\ip{\Phi}{\Psi}$ associating to each pair of vectors
a complex number, and satisfying the following properties: 
\begin{enumerate}[label=(\Alph{enumi})]
\item \label{enu:inner-product-complex-conjugate}$\ip{\Phi}{\Psi}$ is
the complex conjugate of $\ip{\Psi}{\Phi}$;
\item \label{enu:inner-product-linear}$\ip{\Phi}{\Psi}$ is conjugate linear
in its first argument and linear in its second argument; 
\item \label{enu:inner-product-positive-definite}$\ip{\Psi}{\Psi}$ is
always non-negative and is equal to $0$ only if $\ket{\Psi}$ is
the zero vector. 
\end{enumerate}
It turns out that a vector space defined over a finite field cannot
have an inner product satisfying the properties above. However, we
will introduce an Hermitian ``dot product'' satisfying some of those
properties.

@@

\begin{definition}[Hermitian dot product] Given vectors $\ket{\Phi}$
and $\ket{\Psi}\in\mathcal{H}$, it can be shown~\cite{grove2002classical,Wan2006}
the Hermitian dot product is always reducible to the form 
\begin{equation}
\ip{\Phi}{\Psi}=\sum_{i=0}^{D-1}\beta_{i}^{p}\alpha_{i}\,.\label{innerprod}
\end{equation}
If we use $\bra{\Phi}$ to represent the dual vector of $\ket{\Phi}$,
i.e., the conjugate transport of $\ket{\Phi}$, then the matrix multiplication
of $\bra{\Phi}$ and $\ket{\Phi}$ results a $1\times1$ matrix whose
only entry is their Hermitian dot product $\ip{\Phi}{\Psi}$.\end{definition}
